Morphology-based Entity and Relational Entity Extraction Framework for Arabic

### 2018-09-07 tex/merf/TALMajor/tool.tex
final version
applied recommended edits; almost done

Up to our knowledge, 
$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
where $2^{\GG}$ denotes the power set of $\GG$ which is the set of all subsets 
of $\GG$.
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

\begin{figure}[tb]
\setcode{utf8}
\begin{center}
  \resizebox{0.75\columnwidth}{!}{ 
  	{\relsize{-2} \input{figures/syn2ex.tex}} }
\setcode{standard}
\vspace{-3em}
\caption{$Syn^2($\RL{mA'}$)$}
\vspace{-2em}
\label{fig:introsynEx}
\end{center}
\end{figure}

@@

Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
Table~\ref{tab:taskMBF} presents the defined formulae.
The user denotes the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of\_Person}.
The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula $R$ to denote ``relative position'' entities, 
and defines it as a disjunction of terms that check for 
solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
%requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).

@@

The matching parse trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
The first tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
The second tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
%The leaf nodes of the trees are entities and the edges and internal nodes are text, 
%morphology-based, 
%and word distance based relational entities. 
The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.

\vspace{-2em}
\subsection{User-defined relations and actions}
\label{sec:relations-actions}

A relation is defined by the user as a tuple 
$\langle e_1,e_2,r\rangle$ where 
$e_1,e_2,$ and $r$ are identifiers associated with 
sub-expressions of an MRE $f$.
Matches of the relation are a set of labeled binary edges
where matches of $e_1$ and $e_2$ are the source and destination nodes
and matches of $r$ are the edge labels.
We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of the corresponding relation, 
and we refer to them as relational entities.

We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
Let $e_1,o_1,r,o_2,$ and $e_2$ be identifiers to the sub-expressions 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
%We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
Note that there is no match to the optional $O$ formula in match 1.
Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second matching parse tree are \RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.

@@

Moreover, \framework allows advanced users to write C++ code snippets 
to process matches of sub-expressions.
Each sub-expression can be associated with two computational 
actions: \cci{pre-match} and \cci{on-match}.
%
\framework provides an API that enriches the actions with detailed access to
all solution features of an expression or a formula match including 
text, position, length, equivalent numerical value when applicable, 
and morphological features.
The API follows a decorator pattern in that it incrementally adds the 
action results to the matching entities.
Once \framework computes all matching parse trees, it traverses each tree to 
execute the user defined \cci{pre-match} actions in pre-order manner
and the \cci{on-match} actions in post-order manner.
This follows an observer pattern that notifies listeners with each
produced match.

@@

For each MRE, 
\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2012introduction}.
%Each MRE operation has its equivalent representation in an NFA. 
We support the upto operation ($f$\^{}$x$), which is not directly 
supported in~\cite{sipser2012introduction}, by 
%it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
%we can expand it into a standard regular expression form, for example
expanding it into a regular expression form; for example 
$f$\^{}$3$ is equivalent to $f?|ff|fff$. 
Consider the example of Figure~\ref{fig:intromotiv} and the
corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$.
Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
$q_8, q_9, \dots, q_{13}$ represent NFA states,
and edges are transitions based on MBF tags such as 
$P,$ and $N$.
Edges labeled with the empty string $\epsilon$ are non-deterministic.
%The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.

\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
%A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
A simulation match $m$ of an expression $f$ is a parse tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g. Fig.~\ref{fig:taskMRE}.
The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
corresponding to the text sequence 
$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
%
%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
%where $0\le m\le n$, 
%the next simulation starts at $R_{n+1}$. 
%This disallows overlap of matches for the same MRE. 
%
%In case the NFA simulation has no match, 
%the next simulation starts at $R_{m+1}$. 
If we have more than one match for an expression, 
%starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest.
%Figure~\ref{fig:taskMRE} shows two match trees of $\mathit{dir}$ extracted from the text of Figure~\ref{fig:intromotiv}.
%\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$).
%The $+$, $|$, and $?$ MRE operations are internal nodes.

%\framework maintains a function $\phi\subset Q\times\Phi$, 
%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of sub-expressions in the MRE.
%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to sub-expression $f\in \Phi$. 
%\framework uses $\phi$ to compute a match tree with respect to the MRE regular expression. 
%It also uses $\phi$ and the match sequence to compute the sequence of computational actions of an MRE match.

@@

\begin{comment} 
Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding sub-expression.
For instance, we could have multiple matches if we associate $e_i$ with the sub-expression $P|N$ while we have $(P|N)+$.
%to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
%with the smallest nonzero positive distance between the source and the destination
%where the distance is the number of words between the matches. 
Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of sub-expressions:
\begin{itemize}
\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labeled by $r$.
\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labeled by $r$.
\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labeled by $r$. The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labeled by $r$.
\end{itemize}
\end{comment}

%In Figure~\ref{fig:taskMRE}, \framework names the sub-expressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
%The user defines the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.

>>>>

Up to our knowledge, 
$Syn^k$ provides the first light Arabic WordNet based on the lexicon of Sarf.
The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively.
Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
where $2^{\GG}$ denotes the power set of $\GG$ which is the set of all subsets 
of $\GG$.
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

\begin{figure}[tb]
\setcode{utf8}
\begin{center}
  \resizebox{0.75\columnwidth}{!}{ 
  	{\relsize{-2} \input{figures/syn2ex.tex}} }
\setcode{standard}
\vspace{-3em}
\caption{$Syn^2($\RL{mA'}$)$.}
\vspace{-2em}
\label{fig:introsynEx}
\end{center}
\end{figure}

@@

Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
Table~\ref{tab:taskMBF} presents the defined formulae.
The user denotes the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of\_Person}.
The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula $R$ to denote ``relative position'' entities, 
and defines it as a disjunction of terms that check for 
solutions matching stems such as \RL{qrb} (``near'') and \RL{fy} (``in'').
%requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(``first''), \RL{_tAny}(``second''), \dots \RL{`A^sr}(``tenth'').

@@

The matching parse trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
The first tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(``Khalifa Tower next to the first intersection'').
The second tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(``Dubai Mall is located near this building'').
%The leaf nodes of the trees are entities and the edges and internal nodes are text, 
%morphology-based, 
%and word distance based relational entities. 
The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to subexpression matches.
For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the subexpression $(P|N)+$.

\vspace{-1em}
\subsection{User-defined relations and actions}
\label{sec:relations-actions}

A relation is defined by the user as a tuple 
$\langle e_1,e_2,r\rangle$ where 
$e_1,e_2,$ and $r$ are identifiers associated with 
subexpressions of an MRE $f$.
Matches of the relation are a set of labeled binary edges
where matches of $e_1$ and $e_2$ are the source and destination nodes
and matches of $r$ are the edge labels.
We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of the corresponding relation, 
and we refer to them as relational entities.

We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
Let $e_1$, $o_1$, $r$, $o_2,$, and $e_2$ be identifiers to the subexpressions 
$(P|N)+$, $O?$, $R$, $O\wedge 2$, and $(P|N|U)+$, respectively.
%We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (``Khalifa Tower''), \RL{bAlqrb} (``next''), \RL{mn} (``to''), and \RL{AltqA.t` Al-'wl} (``first intersection'').
Note that there is no match to the optional $O$ formula in match 1.
Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second matching tree are \RL{dby mwl} (``Dubai Mall''), \RL{`l_A} (``is located''), \RL{mqrbT} (``near''), \RL{mn h_dA} (``this''), and \RL{Almbn_A} (``building''), respectively.

@@

Moreover, \framework allows advanced users to write C++ code snippets 
to process matches of subexpressions.
Each subexpression can be associated with two computational 
actions: \cci{pre-match} and \cci{on-match}.
%
\framework provides an API that enriches the actions with detailed access to
all solution features of an expression or a formula match including 
text, position, length, equivalent numerical value when applicable, 
and morphological features.
The API follows a decorator pattern in that it incrementally adds the 
action results to the matching entities.
Once \framework computes all matching parse trees, it traverses each tree to 
execute the user defined \cci{pre-match} actions in pre-order manner
and the \cci{on-match} actions in post-order manner.
This follows an observer pattern that notifies listeners with each
produced match.

@@

For each MRE, 
\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2012introduction}.
%Each MRE operation has its equivalent representation in an NFA. 
We support the upto operation ($f$\^{}$x$), which is not directly 
supported in~\citename{sipser2012introduction}~\citeyear{sipser2012introduction}, by 
%it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
%we can expand it into a standard regular expression form, for example
expanding it into a regular expression form; for example 
$f$\^{}$3$ is equivalent to $f?|ff|fff$. 
Consider the example of Figure~\ref{fig:intromotiv} and the
corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$.
Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
$q_8, q_9, \dots, q_{13}$ represent NFA states,
and edges are transitions based on MBF tags such as 
$P,$ and $N$.
Edges labeled with the empty string $\epsilon$ are non-deterministic.
%The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.

\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
%A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
A simulation match $m$ of an expression $f$ is a parse tree where the root spans the expression, the internal nodes are roots to subexpressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g. Figure~\ref{fig:taskMRE}.
The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
corresponding to the text sequence 
$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
%
%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
%where $0\le m\le n$, 
%the next simulation starts at $R_{n+1}$. 
%This disallows overlap of matches for the same MRE. 
%
%In case the NFA simulation has no match, 
%the next simulation starts at $R_{m+1}$. 
If we have more than one match for an expression, 
%starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest.
%Figure~\ref{fig:taskMRE} shows two match trees of $\mathit{dir}$ extracted from the text of Figure~\ref{fig:intromotiv}.
%\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$).
%The $+$, $|$, and $?$ MRE operations are internal nodes.

%\framework maintains a function $\phi\subset Q\times\Phi$, 
%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of subexpressions in the MRE.
%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to subexpression $f\in \Phi$. 
%\framework uses $\phi$ to compute a match tree with respect to the MRE regular expression. 
%It also uses $\phi$ and the match sequence to compute the sequence of computational actions of an MRE match.

@@

\begin{comment} 
Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding subexpression.
For instance, we could have multiple matches if we associate $e_i$ with the subexpression $P|N$ while we have $(P|N)+$.
%to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
%with the smallest nonzero positive distance between the source and the destination
%where the distance is the number of words between the matches. 
Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of subexpressions:
\begin{itemize}
\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labeled by $r$.
\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labeled by $r$.
\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labeled by $r$. The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labeled by $r$.
\end{itemize}
\end{comment}

%In Figure~\ref{fig:taskMRE}, \framework names the subexpressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
%The user defines the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
